
---
title: "Monte Carlo Equity analysis"
Author:  "Dana McDaniel"
Date:  "`r print(Sys.Date())`"
output:
  pdf_document: default

---
```{r, comment="  ", echo = F, warning = F, message = F}
knitr::opts_chunk$set(comment="  ", echo = F, warning = F, message = F)

library(lubridate)
library(quantmod)
library(dplyr)
library(ggplot2)
library(knitr)

```

## Objective

To understand possible stock price performance in order to make better financial decisions.

I was inspired by a recent HR policy change at Genentech.  Employees are now empowered choose one of several mixes of stock SSAR and RSU sets.  Of course, the value of each bundle varies based on the future stock price. This script accepts a stock ticker and some associated parameters as input and uses a log-normal model to estimate ticker price after a specified holding period. This is consistent with the popular, if dated, book by Burton Malkiel:  A Random Walk down Wall Street.  This script does not constitute complete, or even appropriate, financial advice.  Please consult your fiduciary financial advisor for more information.

## Setup and model parameters

```{r, echo = F, warning = F, message = F}

# Modeling details:
stock_ticker <- "ROG.SW"      # IMPORTANT:  which stock do you want to model?
begin_date   <- "2016-06-01"  # IMPORTANT:  capture historical data starting on this date
end_date     <- floor_date(today(), unit = "week") + days(1)  # end on most recent monday
n_days       <- 252 * 1       # IMPORTANT: how many days do you want to model?  252 market days per year
j_prices     <- 750           # IMPORTANT: how many simulations to estimate returns? 
k_sim        <- 7             # how many simulations to visualize?
report_percentiles <- 0:20/20 # what percentile performances do we want to view in table?

```

Equity price estimation for `r stock_ticker` is performed using a log-normal price distribution model.  The price distribution model is constructed from sampling daily ajusted closing price for the stock between `rbegin_date` and `r end_date`.  When I analyze Roche stock, the dates are selected based on when I joined the organization as a contractor.  For other firms, I try to capture time since IPO or the last four years.  Daily price fluctuations are modeled as IID in the distribution range, which is consistent with the "random walk" model of stock performance.  This report visualizes `r k_sim` random walks for information only.  Equity risk analysis is performed by monte carlo simulating `r j_prices` random walks and looking at the distribution of end-of-period holding prices.

```{r, functions}

simulate_closing_price <- function(n_days, price, avg, stdev){
    deltas        <- rnorm(n_days, avg, stdev)
    sample_path   <- cumprod(c(price, 1 + deltas))
    closing_price <- sample_path[1 + n_days]
    closing_price    
}


simulate_ticker_distribution <- function(stock_ticker="ROG.SW", begin_date = '2014-06-01', end_date = '2020-01-01', n_days = 504, j_prices = 750, report_percentiles = 0:20/20) {
  
  stock_raw <- getSymbols(stock_ticker, 
                          auto.assign = FALSE, 
                          src = "yahoo",
                          from = begin_date, 
                          to = end_date) # this creates an RHHBY object
  stock_log_returns <- stock_raw %>%   
    Ad() %>% 
    dailyReturn(type = 'log')
  
  # start simulating prices from most recent date
  # Simulation parameters
  
  stock_avg    <- mean(stock_log_returns)
  stock_stdev  <- sd(stock_log_returns) 
  stock_init_price <- stock_raw %>% 
    Ad() %>% 
    as.data.frame() %>% 
    mutate(., date = rownames(.)) %>% 
    filter(date(date) == max(date)) %>%
    select(ends_with("Adjusted")) %>%
    as.numeric()
  
  stock_closing_price <- replicate(j_prices, 
                                   simulate_closing_price(n_days = n_days, 
                                                          price = stock_init_price, 
                                                          avg = stock_avg, 
                                                          stdev = stock_stdev))
  
  stock_closing_price_gains_pc    <- 100 * (stock_closing_price - stock_init_price) / stock_init_price
  stock_closing_price_percentiles <- quantile(stock_closing_price_gains_pc, 
                                              report_percentiles)
  
  tmp <- data.frame(percentile = report_percentiles) %>%
    mutate(length_days = n_days,
           start_price = stock_init_price,
           final_price = quantile(stock_closing_price, report_percentiles),
           price_fold_change = (final_price - start_price) / start_price)
  
  return(tmp)
}
```
#  Simulated performance over multipe time-periods

```{r}
stock_ticker <-"ROG.SW"
begin_date <- '2014-01-01'
end_date     <- floor_date(today(), unit = "week") + days(1)
num_simulations <- 750
report_percentiles <- 0:200/200
hold_length_days <- 252 # num trading days in a year

one_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
two_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 2,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
three_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 3,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
four_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 4,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
five_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 5,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
six_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 6,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
seven_year_hold <- simulate_ticker_distribution(stock_ticker = stock_ticker,
                                              j_prices = num_simulations,
                                              begin_date = begin_date,
                                              n_days = hold_length_days * 7,
                                              end_date = end_date,
                                              report_percentiles = report_percentiles)
df_list <- list(one_year_hold,
                two_year_hold,
                three_year_hold,
                four_year_hold,
                five_year_hold,
                six_year_hold,
                seven_year_hold)

get_percentile <- function(hold_df) {
  # this estimates the probability of loss based on MCMC simulation.
  pos_pctiles <- which(hold_df[, "price_fold_change"] > 0)
  min_pos_pctile <- hold_df[min(pos_pctiles), "percentile"]
  return(min_pos_pctile)
}

loss_prob <- lapply(df_list,
                               FUN = function(x) get_percentile(x)) %>%
  unlist()

data.frame(hold_length_years = 1:length(positive_change_prob),
           prob_of_loss = loss_prob)
```


# manual analysis
```{r, echo = F, warnings = F, error = F}
stock_dat <- getSymbols(stock_ticker, 
                        auto.assign = FALSE, 
                        src = "yahoo",
                        from = begin_date, to = end_date) # this creates an RHHBY object
stock_log_returns <- stock_dat %>%   
    Ad() %>% 
    dailyReturn(type='log')

# start simulating prices from most recent date
# Simulation parameters

stock_avg    <- mean(stock_log_returns)
stock_stdev  <- sd(stock_log_returns) 
price      <- rep(NA,      n_days * k_sim)
simulation <- rep(1:k_sim, n_days) %>% 
    sort()
stock_init_price <- stock_dat %>% 
    Ad() %>% 
    as.data.frame() %>% 
    mutate(., date = rownames(.)) %>% 
    filter(date(date) == max(date)) %>%
    select(ends_with("Adjusted")) %>%
    as.numeric()

# Monte Carlo aka 0-order MM
#  for each simulation trial, calculate a random walk of stock prices
for(sim in 0:(k_sim - 1)) {
    for(i in 2:n_days) {
        if(i == 2)  { price[sim * n_days + i - 1]  <- stock_init_price }
        delta <- rnorm(1, stock_avg, stock_stdev)
        price[sim * n_days + i] <- price[sim * n_days + i - 1] * exp(delta)   
    }
}

stock_closing_price <- replicate(j_prices, 
                                  simulate_closing_price(n_days = n_days, 
                                                         price = stock_init_price, 
                                                         avg = stock_avg, 
                                                         stdev = stock_stdev))

stock_closing_price_gains_pc    <- 100 * (stock_closing_price - stock_init_price) / stock_init_price
stock_closing_price_percentiles <- quantile(stock_closing_price_gains_pc, 
                                      report_percentiles)

mc_stock_prices <- data.frame(day = 1:n_days, 
                              price = price, 
                              trial = simulation)

closing_price_colnum <- stringr::str_which(colnames(stock_dat), "[.]Close")
plot(stock_dat[, closing_price_colnum], 
     main = stock_ticker)
```

## A random walk of \$`r stock_ticker` over `r n_days` days

```{r, echo = F}
mc_stock_prices %>% 
    ggplot(data=., 
           aes(x = day, 
               y = price, 
               color = factor(trial))) + 
    geom_line(alpha = 0.9) + 
    geom_hline(yintercept = stock_init_price, lty = 2, color = "gray") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          panel.grid.minor = element_blank()) + 
    guides(color = F) + 
    labs(title =paste0("$", stock_ticker, " price simulation over ", round(n_days/252, 1), " years"),
         subtitle =paste0("Monte carlo simulation using price activity between\n", 
                          begin_date, " and ", end_date),
         color = "Trial",
         x = "Holding length (days)",
         y = "Price ($)")


```

## Equity Risk Assessment

Equity risk is modeled here as the variability of final adjusted closing price at the end of the holding range.  Stock ticker \$`r stock_ticker` is compared against Vanguard's S&P 500 index fund, \$VOO.  Note that price variances are large in part due to the skew introduced from the log-normal model here.  One measure of risk is average price growth divided by the standard deviation:  \[\frac{\bar{X}}{\sqrt{\frac{1}{N-1}\sum_{i=1}^{N} (X_{i} -\bar{X})}} \].  An investor will always desire a portfolio with higher reward per unit risk.

```{r}
### repeat MC simulation for VOO

# get data
voo_dat <- getSymbols("VOO", 
                        auto.assign = FALSE, 
                        src = "yahoo",
                        from = begin_date, to = end_date)
voo_log_returns <- voo_dat %>%   
    Ad() %>% 
    dailyReturn(type='log')

# model parameters
voo_avg    <- mean(voo_log_returns)
voo_stdev  <- sd(voo_log_returns) 
voo_price      <- rep(NA,      n_days * k_sim)
voo_simulation <- rep(1:k_sim, n_days) %>% 
    sort()
voo_price[1] <- voo_dat %>% 
    Ad() %>% 
    as.data.frame() %>% 
    mutate(., date = rownames(.)) %>% 
    filter(date(date) == max(date)) %>%
    select(ends_with("Adjusted")) %>%
    as.numeric()

# modeling
voo_closing_price <- replicate(j_prices, 
                               simulate_closing_price(n_days = n_days, 
                                                      price = voo_price[1], 
                                                      avg = voo_avg, 
                                                      stdev = voo_stdev))
voo_closing_price_gains_pc <- 100 * (voo_closing_price - voo_price[1]) / voo_price[1]
voo_closing_price_percentiles <- quantile(voo_closing_price_gains_pc, report_percentiles)

# report out
price_results <- data.frame(percentile = names(stock_closing_price_percentiles),
           stock_price = quantile(stock_closing_price,
                                 report_percentiles),
          stock_closing_price_percentiles,
           voo_closing_price_percentiles) %>%
    mutate(stock_price = round(stock_price, 2),
           stock_closing_price_percentiles = round(stock_closing_price_percentiles, 1),
           voo_closing_price_percentiles = round(voo_closing_price_percentiles, 1)) 
price_results %>% kable(col.names = c("Percentile", 
                        paste0(stock_ticker, " Final Price ($)"),
                        paste0(stock_ticker, " Price change (%)"),
                        paste0("VOO Price change (%)")))



```


`r print(stock_ticker)` reward/risk ratio: `r round(stock_avg/stock_stdev, 3)` vs. VOO reward/risk ratio: `r round(voo_avg/voo_stdev, 3)`

Final prices were monte-carlo simulated from `j_prices` trials using the log-normal model discussed above.  Simple returns on investment (%) are calculated, and a comparison for similar simulation with \$VOO.  Each row in the table below represents the percentile rank performance of \$`r print(stock_ticker)` in monte carlo simulations.

```{r}

stock_closing_price_gains_pc %>% data.frame(price = .) %>%
    ggplot(aes(x = price)) +
        geom_density() +
    geom_density(data = data.frame(price = voo_closing_price_gains_pc), 
                 aes(x = price), color = "gray") +
    geom_vline(xintercept = 0) +
    theme_bw() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5)) +
    labs(title= paste0("Price distribution for $", stock_ticker, " (black)"), 
         subtitle = "Price distribution for $VOO (gray)",
         y = "Probability density",
         x = "Price performance (%)")
```

# Predict value of options

Roche offers options per the Roche S-Sar Plan, which can be looked up at YourLifeYourRewards.com somewhere in the LTI menu offerings.  The SSARs are executed to provide stock per the formula:

G = N x (PE – SP) ÷ PE
Where:
G = the number of Genussscheine to be delivered to the Participant, rounded to
a whole number of Genussscheine
N = total number of S-SARs exercised
PE = the Price at Exercise
SP = the Strike Price

Strike price is defined as "The Strike Price of a particular S-SAR shall be stated in the S-SAR Grant Notice and shall be not less than the Market Value of a Genussschein on the Grant Date"

Notice that Roche performs rounding to the nearest whole number.  There's an exploit to the SSAR system: the number of Genusscheine are given to you rounded to the nearest integer.  You want to transact the minimum number of SSARs that grant a Genusscheine

```{r}
sp <- c(251.90, 220.80) # 2017 and 2018 grants
pe <- Cl(stock_dat)[end_date] %>% as.numeric()

g <- (pe - sp) / pe

transact_min[1] <- min(which(g[1] * 1:20 > 0.51))
transact_min[2] <- min(which(g[2] * 1:20 > 0.51))
transact_min

# todo calculate value of SSAR option based on price and transact_min
```

