
---
title: "Monte Carlo Equity analysis"
Author:  "Dana McDaniel"
Date:  "`r print(Sys.Date())`"
output:
  pdf_document: default

---
```{r, comment="  ", echo = F, warning = F, message = F}
knitr::opts_chunk$set(comment="  ", echo = F, warning = F, message = F)

library(lubridate)
library(quantmod)
library(dplyr)
library(ggplot2)
library(knitr)

```

## Objective

To understand possible stock price performance in order to make better financial decisions.

I was inspired by a recent HR policy change at Genentech.  Employees are now empowered choose one of several mixes of stock SSAR and RSU sets.  Of course, the value of each bundle varies based on the future stock price. This script accepts a stock ticker and some associated parameters as input and uses a log-normal model to estimate ticker price after a specified holding period. This is consistent with the popular, if dated, book by Burton Malkiel:  A Random Walk down Wall Street.  This script does not constitute complete, or even appropriate, financial advice.  Please consult your fiduciary financial advisor for more information.

## Setup and model parameters

```{r, echo = F, warning = F, message = F}

# Modeling details:
stock_ticker <- "ROG.VX"      # IMPORTANT:  which stock do you want to model?
begin_date   <- "2014-06-01"  # IMPORTANT:  capture historical data starting on this date
end_date     <- floor_date(today(), unit = "week") + days(1)  # end on most recent monday
n_days       <- 252           # IMPORTANT: how many days do you want to model?  252 market days per year
j_prices     <- 500           # IMPORTANT: how many simulations to estimate returns? 
k_sim        <- 7             # how many simulations to visualize?

```

Equity price estimation for `r stock_ticker` is performed using a log-normal price distribution model.  The price distribution model is constructed from sampling daily ajusted closing price for the stock between `begin date` and `end date`.  These dates were arbitrarily selected based on when I joined gRED as a bioinformatics associate and scientific software specialist.  Daily price fluctuations are modeled as IID in the distribution range, which is consistent with the "random walk" model of stock performance.  This report visualizes `k_sim` random walks for information only.  Equity risk analysis is performed by monte carlo simulating `r j_prices` random walks and looking at the distribution of end-of-period holding prices.

```{r, echo = F, warnings = F, error = F}
stock_dat <- getSymbols(stock_ticker, 
                        auto.assign = FALSE, 
                        src = "yahoo",
                        from = begin_date, to = end_date) # this creates an RHHBY object
stock_log_returns <- stock_dat %>%   
    Ad() %>% 
    dailyReturn(type='log')

# start simulating prices from most recent date
# Simulation parameters

stock_avg    <- mean(stock_log_returns)
stock_stdev  <- sd(stock_log_returns) 
price      <- rep(NA,      n_days * k_sim)
simulation <- rep(1:k_sim, n_days) %>% 
    sort()
stock_init_price <- stock_dat %>% 
    Ad() %>% 
    as.data.frame() %>% 
    mutate(., date = rownames(.)) %>% 
    filter(date(date) == max(date)) %>%
    select(ends_with("Adjusted")) %>%
    as.numeric()

# Monte Carlo aka 0-order MM
#  for each simulation trial, calculate a random walk of stock prices
for(sim in 0:(k_sim - 1)) {
    for(i in 2:n_days) {
        if(i == 2)  { price[sim * n_days + i - 1]  <- stock_init_price }
        delta <- rnorm(1, stock_avg, stock_stdev)
        price[sim * n_days + i] <- price[sim * n_days + i - 1] * exp(delta)   
    }
}


simulate_closing_price <- function(n_days, price, avg, stdev){
    deltas        <- rnorm(n_days, avg, stdev)
    sample_path   <- cumprod(c(price, 1 + deltas))
    closing_price <- sample_path[1 + n_days]
    closing_price    
}

stock_closing_price <- replicate(j_prices, 
                                  simulate_closing_price(n_days = n_days, 
                                                         price = stock_init_price, 
                                                         avg = stock_avg, 
                                                         stdev = stock_stdev))

stock_closing_price_gains_pc    <- 100 * (stock_closing_price - stock_init_price) / stock_init_price
stock_closing_price_percentiles <- quantile(stock_closing_price_gains_pc, 
                                      c(0.1, .25, 0.33, 0.5, 0.66,  0.75, 0.9))

mc_stock_prices <- data.frame(day = 1:n_days, 
                              price = price, 
                              trial = simulation)


```

## A random walk of \$`r stock_ticker` over `r n_days` days

```{r, echo = F}
mc_stock_prices %>% 
    ggplot(data=., 
           aes(x = day, 
               y = price, 
               color = factor(trial))) + 
    geom_line(alpha = 0.9) + 
    geom_hline(yintercept = stock_init_price, lty = 2, color = "gray") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5)) + 
    guides(color = F) + 
    labs(title =paste0("$", stock_ticker, " price simulation"),
         subtitle =paste0("Monte carlo simulation using price activity between\n", 
                          begin_date, " and ", end_date),
         color = "Trial",
         x = "Holding length (days)",
         y = "Price ($)")


```

## Equity Risk Assessment

Equity risk is modeled here as the variability of final adjusted closing price at the end of the holding range.  Stock ticker \$`r stock_ticker` is compared against Vanguard's S&P 500 index fund, \$VOO.  Note that price variances are large in part due to the skew introduced from the log-normal model here.  One measure of risk is average price growth divided by the standard deviation:  \[\frac{\bar{X}}{\sqrt{\frac{1}{N-1}\sum_{i=1}^{N} (X_{i} -\bar{X})}} \].  An investor will always desire a portfolio with higher reward per unit risk.

```{r}
### repeat MC simulation for VOO

# get data
voo_dat <- getSymbols("VOO", 
                        auto.assign = FALSE, 
                        src = "yahoo",
                        from = begin_date, to = end_date)
voo_log_returns <- voo_dat %>%   
    Ad() %>% 
    dailyReturn(type='log')

# model parameters
voo_avg    <- mean(voo_log_returns)
voo_stdev  <- sd(voo_log_returns) 
voo_price      <- rep(NA,      n_days * k_sim)
voo_simulation <- rep(1:k_sim, n_days) %>% 
    sort()
voo_price[1] <- voo_dat %>% 
    Ad() %>% 
    as.data.frame() %>% 
    mutate(., date = rownames(.)) %>% 
    filter(date(date) == max(date)) %>%
    select(ends_with("Adjusted")) %>%
    as.numeric()

# modeling
voo_closing_price <- replicate(j_prices, 
                               simulate_closing_price(n_days = n_days, 
                                                      price = voo_price[1], 
                                                      avg = voo_avg, 
                                                      stdev = voo_stdev))
voo_closing_price_gains_pc <- 100 * (voo_closing_price - voo_price[1]) / voo_price[1]
voo_closing_price_percentiles <- quantile(voo_closing_price_gains_pc, 
                                      c(0.1, .25, 0.33, 0.5, 0.66,  0.75, 0.9))

# report out
price_results <- data.frame(percentile = names(stock_closing_price_percentiles),
           stock_price = quantile(stock_closing_price,
                                 c(0.1, .25, 0.33, 0.5, 0.66,  0.75, 0.9)),
          stock_closing_price_percentiles,
           voo_closing_price_percentiles) %>%
    mutate(stock_price = round(stock_price, 2),
           stock_closing_price_percentiles = round(stock_closing_price_percentiles, 1),
           voo_closing_price_percentiles = round(voo_closing_price_percentiles, 1)) 
price_results %>% kable(col.names = c("Percentile", 
                        paste0(stock_ticker, " Final Price ($)"),
                        paste0(stock_ticker, " Price change (%)"),
                        paste0("VOO Price change (%)")))
```


`r print(stock_ticker)` reward/risk ratio: `r round(stock_avg/stock_stdev, 3)` vs. VOO reward/risk ratio: `r round(voo_avg/voo_stdev, 3)`

Final prices were monte-carlo simulated from `j_prices` trials using the log-normal model discussed above.  Simple returns on investment (%) are calculated, and a comparison for similar simulation with \$VOO.  Each row in the table below represents the percentile rank performance of \$`r print(stock_ticker)` in monte carlo simulations.

```{r}

stock_closing_price_gains_pc %>% data.frame(price = .) %>%
    ggplot(aes(x = price)) +
        geom_density() +
    geom_density(data = data.frame(price = voo_closing_price_gains_pc), 
                 aes(x = price), color = "gray") +
    geom_vline(xintercept = 0) +
    theme_bw() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5)) +
    labs(title= paste0("Price distribution for $", stock_ticker, " (black)"), 
         subtitle = "Price distribution for $VOO (gray)",
         y = "Probability density",
         x = "Price performance (%)")
```

